%% MS-ML Studio / MLMS Studio — Elsevier preprint (elsarticle)
%% Sincronizado com documentation/Article_Improved.md (Mar 2026).

\documentclass[preprint,12pt]{elsarticle}
\usepackage{mathtools}
\usepackage{float}
\usepackage{amssymb,natbib}
\usepackage{booktabs}

\journal{Knowledge Based Systems}

\begin{document}

\begin{frontmatter}

\title{MS-ML Studio: An open-source MATLAB tool for unified mass-spectrometry preprocessing, feature extraction, and classification}

\address[addr1]{Departamento de Bioinform{\'a}tica, Universidade Federal do Paran{\'a}, Brazil}
\address[addr2]{Setor Litoral, Universidade Federal do Paran{\'a}, Brazil}
\address[addr3]{Departamento de Patologia M{\'e}dica, Universidade Federal do Paran{\'a}, Brazil}

\author[addr1]{Calebe Elias Ribeiro Brima}
\author[addr1]{Marilson Reque}
\author[addr3]{Cibelle B. Dallagassa}
\author[addr3]{Cyntia M. T. Fadel-Picheth}
\author[addr2]{Luciano F. Huergo}
\author[addr1]{Roberto Tadeu Raittz\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{raittz@ufpr.br}

\begin{abstract}
Although many mass-spectrometry (MS) preprocessing and peak-detection tools exist, few free, open-source options integrate raw-spectral import, discriminant feature learning, and classifier training in one interactive graphical workflow for users without programming expertise. \textbf{MS-ML Studio} is an open-source MATLAB application that unifies preprocessing (normalization, reshaping, optional alignment), three complementary, cross-validated feature extractors (correlation-based peak selection, Watch Points, Feature Resonance with genetic-algorithm optimization), and multilayer perceptron (MLP) classification with in-GUI visualization. As a \textbf{reference proof-of-concept}---not a clinical validation---we applied the pipeline to intact-cell MALDI-TOF spectra of four \textit{Aeromonas} species in one-against-all tasks; held-out test accuracies ranged from 82.1\% to 100\% (F1-scores 0.769--1.000), with the smallest test partition (\textit{A.~veronii}/\textit{sobria}, $n=30$) warranting cautious interpretation. Source code and executable are distributed under an open-source license (repository URL and Zenodo DOI to be inserted before submission), with a versioned archive for long-term citation in line with journal requirements.
\end{abstract}

\begin{keyword}
mass spectrometry \sep machine learning \sep feature extraction \sep peak selection \sep biomarker discovery \sep MALDI-TOF \sep bacterial identification \sep open-source software
\end{keyword}

\end{frontmatter}

\section{Introduction}

\subsection{Mass spectrometry in biomedical research}
Mass spectrometry (MS) measures mass-to-charge ratios of ionized analytes and underpins proteomics, metabolomics, and microbial identification workflows~\cite{Arnold2006}. Matrix-assisted laser desorption/ionization time-of-flight (MALDI-TOF) MS is now widely used for rapid bacterial identification from intact-cell fingerprints~\cite{Krishnamurthy1996,Croxatto2012}. Across domains, comparable computational stages recur: import, preprocessing, feature extraction, and supervised or statistical classification---steps whose quality strongly affects downstream accuracy~\cite{Wagner2003,Yang2009}. The present work focuses on \textbf{software} that makes this full chain accessible in one graphical environment rather than on new MS instrumentation or large-scale clinical benchmarking.

\subsection{Existing software landscape and its limitations}
A variety of software tools address subsets of the MS analysis pipeline. Instrument-specific proprietary solutions---such as workflows tied to Ciphergen ProteinChip technology~\cite{Conrads2004} and MALDI Biotyper (Bruker Daltonics)~\cite{Buchan2012}---provide integrated workflows but restrict users to vendor-specific data formats and algorithms. Platform-independent open-source tools, including MZmine~2~\cite{Pluskal2010}, OpenMS, Cromwell, and XCMS~\cite{Yang2009}, offer flexible preprocessing algorithms encompassing smoothing, baseline correction, and peak detection. However, these tools generally do not incorporate GA-based or correlation-driven feature extraction coupled to machine learning classifiers in a unified graphical interface accessible to laboratory biologists without programming expertise.

Feature extraction from MS data poses challenges distinct from general dimensionality reduction: spectral peaks must be aligned across samples with varying m/z resolution, informative peaks must be distinguished from noise and non-discriminant background, and the resulting feature representation must generalize across training and validation partitions to avoid overfitting. Genetic-algorithm (GA)-based approaches to biomarker discovery have shown promise~\cite{Armananzas2011,Liu2013}, as has the use of cross-validation within the feature extraction loop~\cite{Liu2013}. Nevertheless, implementations are typically distributed as standalone scripts or require custom integration, presenting a significant barrier to adoption.

\subsection{Contribution of this work}
We developed MS-ML Studio to address the gap between instrument-specific proprietary suites and fragmented open-source components. Our contributions are as follows:
\begin{enumerate}
  \item \textbf{Unified pipeline}: A single MATLAB graphical application that covers all steps from raw spectral import through MLP-based classification and result visualization.
  \item \textbf{Novel feature extraction strategies}: Three complementary methods---correlation-based Best Peaks Selection, Watch Points with fuzzy triangular weighting, and Feature Resonance using GA-optimized multi-sinusoidal functions---each guided by K-fold cross-validation to control overfitting.
  \item \textbf{Platform independence}: The tool handles raw two-column text spectra from any MS platform, without dependence on proprietary data formats.
  \item \textbf{Illustrative performance}: On the reference intact-cell MALDI-TOF \textit{Aeromonas} case study (Sections~\ref{sec:data}--\ref{sec:results}), test-set accuracies of 82--100\% show that the pipeline can perform competitively under transparent settings; features remain exportable to external classifiers via WEKA.
  \item \textbf{Open source}: Full source code and compiled executable are publicly available, enabling reproducibility and extension by the community.
\end{enumerate}

\textit{Note (branding):} \textbf{MS-ML Studio} denotes the software described in this manuscript; the version-controlled distribution is maintained under the \textbf{MLMS Studio} project name in the public repository (see Availability).

\section{Materials and methods}
\label{sec:methods}

\subsection{Data}
\label{sec:data}
The following \textbf{reference case study} illustrates MS-ML Studio on published intact-cell MALDI-TOF data; it is intended as \textbf{proof-of-concept software validation}, not as a standalone clinical or epidemiological study. All mass spectra were acquired on a Bruker MALDI-TOF analyzer from intact bacterial cells of four \textit{Aeromonas} species: \textit{A.~hydrophila}, \textit{A.~caviae}, \textit{A.~veronii}/\textit{sobria}, and \textit{A.~trota}. Sample preparation, bacterial culture, and original spectral acquisition followed the protocol described by Surek \textit{et al.}~\cite{Surek2010}. A total of 567 spectra were available across the four species (316 for the \textit{A.~veronii}/\textit{sobria} group). For each of four one-against-all classification experiments, 90\% of the available spectra were reserved for feature extraction and classifier training; the remaining 10\% were withheld as a blind test set (Table~\ref{tab:partition}).

\subsection{MS-ML Studio architecture}
MS-ML Studio comprises two components: the MS-ML Library, a set of modular MATLAB functions implementing all algorithmic steps, and the MS-ML User Interface, a MATLAB GUIDE-based graphical application that exposes library functionality through interactive controls. The overall workflow is organized into three phases: (1)~preprocessing, (2)~feature extraction with cross-validation, and (3)~neural network training and classification.

\subsection{Preprocessing}

\subsubsection{Data import}
Input spectra are plain-text files with two whitespace-delimited columns: m/z values and relative intensities. This format is instrument-agnostic and can be exported from most MS platforms. There is no built-in converter for proprietary formats such as mzML or mzXML; conversion must be performed with external tools prior to import.

\subsubsection{Intensity normalization}
Intensity normalization rescales each spectrum's intensity vector to the range $[0,1]$ using the min--max formula
\begin{equation}
N(x, x_{\max}, x_{\min}) = \frac{x_i - x_{\min}}{x_{\max} - x_{\min}}\,,
\label{eq:norm}
\end{equation}
where $x$ is the intensity vector and $i$ indexes each data point.

\subsubsection{Spectral reshaping}
High-resolution spectra may contain tens of thousands of data points per spectrum. The reshaping operation maps each spectrum onto a user-defined grid of $W$ evenly spaced bins spanning the global m/z range:
\begin{equation}
G(x; W, x_{\max}, x_{\min}) = \mathrm{round}\left(W \cdot N(x; x_{\max}, x_{\min})\right)\,.
\label{eq:reshape}
\end{equation}
For each bin $i \in \{1,\ldots,W\}$, the output intensity is the maximum of all input intensities whose rounded grid index equals $i$, and the output m/z is the mean of the corresponding input m/z values. Reshaping achieves dimensionality reduction and coarse alignment of peak positions across spectra.

\subsubsection{Peak alignment}
Spectral alignment is applied after reshaping using the Icoshift algorithm~\cite{Tomasi2011}, which employs gap insertion to maximize the correlation between spectra at user-selected m/z windows. Alignment is optional when internal calibration is sufficient for peak reproducibility across runs.

\subsubsection{Peak selection via fuzzy moving average}
MS-ML Studio implements a Fuzzy Moving Average (FMA) detector that computes, for each spectral position $p$, a weighted local average
\begin{equation}
S(x; p, w) = \frac{\sum_{i=p-w}^{p+w} x_i \cdot T(2w+1, i)}{2w+1}\,,\quad p \in \mathbb{N}^{*}\mid 1 \leq p-w,\; p+w \leq S\,,
\label{eq:fma}
\end{equation}
where $w$ is the half-width of the triangular window, $x$ is the normalized intensity vector, and $T(L,i)$ is the triangular weighting function defined in Section~\ref{sec:watchpoints}. Intensity values exceeding $S(x;p,w)$ at local maxima are retained as candidate peaks.

\subsection{Best peaks selection}
From the set of candidate peaks identified by FMA, Best Peaks Selection identifies the subset most discriminant for classification. \textbf{Correlation-based selection}: for each candidate peak position $j$, the Pearson correlation between the presence/absence vector of peak $j$ across training spectra and the binary class label vector is computed; peaks whose absolute correlation exceeds a threshold $\tau$ are retained. \textbf{Probability-based selection}: discriminancy and exclusivity scores are computed from class-conditional peak presence probabilities as in the Markdown manuscript. Both methods produce a ranked list of discriminant spectral positions for subsequent feature extraction.

\subsection{Feature extraction}
\label{sec:watchpoints}
To avoid overfitting, feature extraction algorithms embed K-fold cross-validation within the optimization loop: at each GA iteration, the training set is split into training and validation folds; a configuration is accepted only if the mean of training and validation correlations improves.

\subsubsection{Watch Points}
Each watch point $k$ centered at position $c_k$ uses a triangular weighting window of half-width $L_k$. For $L$ odd,
\begin{equation}
w(L,n) = \begin{cases} \dfrac{2n}{L+1} & 1 \leq n \leq \dfrac{L+1}{2} \\[\medskipamount]
2 - \dfrac{2n}{L+1} & \dfrac{L+1}{2}+1 \leq n \leq L \end{cases}\,.
\end{equation}
For $L$ even,
\begin{equation}
w(L,n) = \begin{cases} \dfrac{2n-1}{L} & 1 \leq n \leq \dfrac{L}{2} \\[\medskipamount]
2 - \dfrac{2n-1}{L} & \dfrac{L}{2}+1 \leq n \leq L \end{cases}\,.
\end{equation}
The feature value for watch point $k$ on spectrum $s$ is $W_k = \sum w_k \cdot p$, where $p$ is the binary peak-presence vector padded to match $w_k$. The GA optimizes window half-widths $\{L_k\}$ to maximize the mean absolute Pearson correlation between $\{W_k\}$ and class labels.

\subsubsection{Feature Resonance}
Feature Resonance encodes peak information as the projection onto a sum of $M$ sinusoidal waves with parameters optimized by the GA:
\begin{equation}
S(x; \mathbf{a}, \mathbf{b}, \mathbf{c}) = \sum_{i=1}^{M} a_i \sin(b_i + c_i x)\,.
\label{eq:resonance}
\end{equation}
The fitness function is the absolute mean Pearson correlation between projection values and class labels. Variants process selected m/z values, best-peak intensities, best-peak m/z values, or watch-point feature vectors.

\subsection{Classification}
Extracted features from all enabled extractors are concatenated and passed to a multilayer perceptron (MLP) implemented in the MS-ML Library. The MLP is trained using MATLAB's Neural Network Toolbox with Levenberg--Marquardt backpropagation. The training/test split follows the 90\%/10\% per-species partition. Features can be exported in ARFF format for WEKA~\cite{Powers2011}.

\subsection{Performance metrics}
Classification performance was evaluated using accuracy, precision, recall, and F1-score~\cite{Powers2011}:
\begin{align}
\mathrm{Accuracy} &= \frac{TP + TN}{N}\,,\quad
\mathrm{Precision} = \frac{TP}{TP + FP}\,,\quad
\mathrm{Recall} = \frac{TP}{TP + FN}\,, \\
\mathrm{F1} &= \frac{2 \cdot \mathrm{Precision} \cdot \mathrm{Recall}}{\mathrm{Precision} + \mathrm{Recall}}\,.
\end{align}

\section{Results}
\label{sec:results}

\subsection{Software overview}
MS-ML Studio comprises the MS-ML Library and an interactive GUI divided into panels for Preprocessing, Peaks Selection, Feature Extraction, and Classification, with real-time visualization of preprocessing, optimization progress, and classification outputs (confusion matrices, PCA scatter plots).

\subsection{Experimental setup (reference case study)}
Four one-against-all binary problems were formulated (one per species). Table~\ref{tab:partition} summarizes dataset partitioning.

\begin{table}[H]
  \centering
  \caption{Dataset partitioning for each one-against-all classification experiment.}
  \label{tab:partition}
  \begin{tabular}{@{}lccc@{}}
    \toprule
    Species & Training ($n$) & Test ($n$) & Total ($n$) \\
    \midrule
    \textit{A.~hydrophila} & 511 & 56 & 567 \\
    \textit{A.~caviae} & 511 & 56 & 567 \\
    \textit{A.~veronii}/\textit{sobria} & 286 & 30 & 316 \\
    \textit{A.~trota} & 511 & 56 & 567 \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Classification performance}
Table~\ref{tab:performance} summarizes metrics on training and test sets. Test F1-scores ranged from 0.769 (\textit{A.~veronii}/\textit{sobria}) to 1.000 (\textit{A.~trota}).

\begin{table}[H]
  \centering
  \caption{Classification performance for each \textit{Aeromonas} one-against-all experiment.}
  \label{tab:performance}
  \small
  \begin{tabular}{@{}llcccc@{}}
    \toprule
    Species & Set & Recall & Precision & Accuracy & F1 \\
    \midrule
    \textit{A.~hydrophila} & Training & 0.889 & 0.956 & 0.898 & 0.921 \\
    & Test & 0.895 & 0.971 & 0.911 & 0.931 \\
    \textit{A.~caviae} & Training & 0.902 & 0.839 & 0.843 & 0.869 \\
    & Test & 0.838 & 0.886 & 0.821 & 0.861 \\
    \textit{A.~veronii}/\textit{sobria} & Training & 0.952 & 0.819 & 0.944 & 0.881 \\
    & Test & 0.833 & 0.714 & 0.900 & 0.769 \\
    \textit{A.~trota} & Training & 1.000 & 0.762 & 0.990 & 0.865 \\
    & Test & 1.000 & 1.000 & 1.000 & 1.000 \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Feature extraction configurations}
Configurations matched the improved manuscript: combined extractors for each species (FR Selected M/Z, FR Best Peaks Intensity/M/Z, Watch Points, multiple correlation thresholds). PCA of extracted features showed clearest separation for \textit{A.~trota}, consistent with perfect test performance; overlap remained visible for \textit{A.~veronii}/\textit{sobria} and \textit{A.~caviae}, reflecting higher phenotypic similarity within the genus.

\section{Discussion}

MS-ML Studio achieved test-set accuracies from 82.1\% (\textit{A.~caviae}) to 100\% (\textit{A.~trota}) and F1-scores from 0.769 to 1.000, consistent with comparable MALDI-TOF studies~\cite{Bohme2010,Croxatto2012}. Variability across species aligns with phylogenetic relatedness; the small test set for \textit{A.~veronii}/\textit{sobria} ($n=30$) limits precision of that point estimate.

The three feature strategies (Best Peaks Selection, Watch Points, Feature Resonance) are complementary; ensemble use generally outperforms a single extractor~\cite{Liu2013}. K-fold cross-validation inside the GA mitigates overfitting. Compared with tools such as MZmine~2~\cite{Pluskal2010}, MS-ML Studio couples preprocessing, discriminant feature learning, and MLP training in one session; WEKA export supports external classifiers.

\textbf{Limitations} include dataset size, plain-text input only (no native mzML/mzXML), MATLAB or MATLAB Runtime dependency, single embedded classifier without integrated significance testing---as detailed in the Markdown source. \textbf{Future work} includes standard format parsers, wavelet-based features~\cite{Liu2013}, unsupervised visualization, and multi-class formulations.

\section{Conclusion}
We presented MS-ML Studio, an open-source MATLAB platform integrating MS preprocessing, three cross-validated feature extractors, and MLP classification in one GUI. On a reference \textit{Aeromonas} MALDI-TOF task, held-out test accuracies of 82--100\% and F1-scores of 0.769--1.000 illustrate competitive performance under transparent settings; larger, multi-site studies are required for clinical claims. Modular library design and WEKA export support extension and broader ML workflows.

\section*{Availability}
\textbf{Software:} Source and executable for MS-ML Studio (MLMS Studio) will be released under an open-source license approved by the institution; replace repository URL and Zenodo DOI placeholders before submission.

\textbf{Data:} The \textit{Aeromonas} MALDI-TOF spectra used in this case study derive from the dataset described by Surek \textit{et al.}~\cite{Surek2010} and are available upon request from the corresponding author.

\section*{Conflict of interest}
The authors declare no conflict of interest.

\section*{Acknowledgements}
The authors thank M.~Surek for providing the \textit{Aeromonas} MALDI-TOF dataset. This work was supported by Universidade Federal do Paran{\'a}.

\bibliographystyle{elsarticle-num}
\bibliography{mendeley,msml_supp}

\end{document}
